Natural selection in contemporary humans is linked to income and substitution effects

### 2021-01-12 why-natural-selection.Rmd
Final tweaks

---
title: "Natural selection in contemporary humans is linked to income and substitution effects"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among lower-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Studying the
mechanisms behind natural selection can help us better understand how individual
differences in complex traits and disease risk arise. Recent work confirms that
natural selection is taking place in modern human populations, using genome-wide
analysis
[@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
particular, genetic variants associated with higher educational attainment are
being selected against, although effect sizes appear small.

@@

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals, corrected for ancestry using 100 genetic principal components (see
Materials and Methods). Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores in the sample by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order of 5% of
a standard deviation. These changes could reflect natural selection within the
UK population, but also emigration, or ascertainment bias within the sample.
Respondents have higher income and are better educated than the UK population,
and they may also differ on other unobserved characteristics
[@10.1093/aje/kwx246]. Since richer and educated people also live longer, this
bias could also increase with age.

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with earnings in a respondent's first
job, and against its correlation with educational attainment, a
predictor of lifetime earnings.^[animation] The raw relationships (left column) are strongly
negative. If we plot selection effects controlling for age at first live birth
(right column), the effect reverses and becomes positive. Thus, the labour
market appears to play an important role in natural selection. Substitution
effects dominate income effects overall, which fits the known association
between income and lower fertility [@becker1960economic;@jones2006economic].

^[animation]: See also the supplementary animation at https://xxx.xxx.xxx

@@

Polygenic scores which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to disease risk, or are what many people would see
as undesirable to have. For example, most people would probably prefer to have
high educational attainment, a low risk of major depressive disorder, and a low
risk of coronary artery disease, but natural selection is pushing against
genes associated with these traits. Potentially, this could increase the health
burden on modern populations, but that depends on effect sizes. Our results
suggest that naïve estimates can be affected by sample ascertainment bias. This
problem may be less serious in surveys which aim to be representative (as the UK
Biobank does not). However, there is still scope for bias, since not all
respondents consent to the collection of genetic data. For instance, completion
rates for genotype data in the US Health and Retirement Study were around 80-85%
[@hrs2020]. Researchers should be aware of the risks of ascertainment when
studying modern natural selection.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases inequality for 29 out of 33 polygenic
scores (Appendix Table \@ref(tab:tbl-inequality-effects)). Since many polygenic
scores are predictive of disease risk, this increase in inequality could
potentially increase health inequalities. In general, the demographic history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may help in the design
of health policies.

>>>>

---
title: "Natural selection in contemporary humans is linked to income and substitution effects"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. Studying these can help us understand how the genetic architecture of health outcomes arise. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Studying the
mechanisms behind natural selection can help us better understand how individual
differences in complex traits and disease risk arise [@benton2021influence]. Recent work confirms that
natural selection is taking place in modern human populations, using genome-wide
analysis
[@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
particular, genetic variants associated with higher educational attainment are
being selected against, although effect sizes appear small.

@@

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals, corrected for ancestry using 100 genetic principal components (see
Materials and Methods). Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores in the sample by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order of 5% of
a standard deviation. These changes could reflect natural selection within the
UK population, but also emigration, or ascertainment bias within the sample.
Respondents have higher income and are better educated than the UK population,
and they may also differ on other unobserved characteristics
[@10.1093/aje/kwx246]. Since richer and more educated people also live longer, this
bias could also increase with age.

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with earnings in a
respondent's first job, and against its correlation with educational attainment,
a predictor of lifetime earnings. (See also the Supplementary Animations for the
uncorrected effects per AFLB group, which show the correlation reversing after
age 22.) The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. Thus, the labour market appears to play an
important role in natural selection. Substitution effects dominate income
effects overall, which fits the known association between income and lower
fertility [@becker1960economic;@jones2006economic].

@@

Polygenic scores which correlate with high (low) earnings and more (less)
education are being selected against (selected for). In addition, many of the
phenotypes under positive selection are linked to disease risk, or are what many
people would see as undesirable to have. For example, most people would probably
prefer to have high educational attainment, a low risk of ADHD and major
depressive disorder, and a low risk of coronary artery disease, but natural
selection is pushing against genes associated with these traits. Potentially,
this could increase the health burden on modern populations, but that depends on
effect sizes. Our results suggest that naïve estimates can be affected by sample
ascertainment bias. This problem may be less serious in surveys which aim to be
representative (as the UK Biobank does not). However, there is still scope for
bias, since not all respondents consent to the collection of genetic data. For
instance, completion rates for genotype data in the US Health and Retirement
Study were around 80-85% [@hrs2020]. Researchers should be aware of the risks of
ascertainment when studying modern natural selection.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases inequality for 29 out of 33 polygenic
scores (Appendix Table \@ref(tab:tbl-inequality-effects)). Since many polygenic
scores are predictive of disease risk, this increase in inequality could
potentially increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.
